---
title: "Lecture-05 Examples"
author: "Christopher Prener, Ph.D."
date: '(`r format(Sys.time(), "%B %d, %Y")`)'
output: 
  github_document: default
  html_notebook: default 
---

## Introduction
This notebook covers basic map production in `R` using a variety of tools. We'll review some of what we discussed during [Lecture-01](https://slu-soc5650.github.io/docs/lecture-01/) and then introduce two new approaches for mapping.

## Dependencies
This notebook requires a variety of packages for working with spatial data:

```{r load-packages}
# spatial packages
library(leaflet)      # interactive maps
library(mapview)      # preview spatial data
library(sf)           # spatial data tools

# other packages
library(here)         # file path management
library(RColorBrewer) # color palettes
```

For best results, you should also have PhantomJS installed as well:

```{r install-phantom-js, eval=FALSE}
webshot::install_phantomjs()
```

## Load Data (Lecture-01 Review)
This notebook requires the data stored in `data/example-data/`. Remember that we use `sf::st_read()` to load shapefile data:

```{r load-data}
city <- st_read(here("data", "example-data", "STL_BOUNDARY_City", "STL_BOUNDARY_City.shp"), stringsAsFactors = FALSE)
nhoods <- st_read(here("data", "example-data", "STL_DEMOS_Nhoods", "STL_DEMOS_Nhoods.shp"), stringsAsFactors = FALSE)
```

## Projections
We haven't talked about projections, yet. They are a part of how we render geometric data on two-dimensional maps. ArcGIS Pro is a bit more flexible than `R` packages are when it comes to projections For now, the more important thing to know is that we need to adjust the projection being used for our data. We'll use two projections today:

* WGS 1984 (crs = `4326`)
* UTM 15N (crs = `32615`)

To ensure that our data are projected correctly, we use `sf::st_transform()`:

```{r project-data}
# city boundary
city <- st_transform(city, crs = 32615)
city84 <- st_transform(city, crs = 4326)

# neighborhood demographics
nhoods <- st_transform(nhoods, crs = 32615)
nhoods84 <- st_transform(nhoods, crs = 4326)
```

## Exploring Data
There are two ways we can explore our data - we can inspect them using functions that print particular properties of each object, and we can inspect them visually. 

### Listing Properties
The `utils::str()` function gives us a chance to preview the data in terms of variable names and the a quick snapshot of the data each variable contains. We can do this for the `city` data:

```{r city-str}
str(city)
```

We can also do this for the `nhoods` data:

```{r nhoods-str}
str(nhoods)
```

### Visually Inspecting Objects

We can also visually inspect the tabular data by clicking on objects in the Environment tab or, alternatively, using `View()` in the console - `View(city)`. To get a preview of the spatial data, we can use the `mapview` package's `mapview()` function. Once the map is created, we can alter the basemap by clicking the icon under the zoom in and out buttons. We can preview the tabular data for each feature by clicking on it:

```{r city-view}
mapview(city)
```

Here is the preview for the `nhoods` data:

```{r nhoods-view}
mapview(nhoods)
```

Getting a sense of how the data are structured, both in terms of their tabular attributes and geometric properties, is the first thing you should do when you bring data into `R`.

## Interactive Mapping with `leaflet` (Lecture-01 Review)
### Basic Mapping of Geometric Objects
During Lecture-01, we discussed the creation of interactive maps using the `leaflet` package. [Leaflet](https://leafletjs.com) is a JavaScript library for creating interactive maps. It is primarily focused on web and mobile mapping. The `leaflet` package for `R` provides access to the JavaScript library. 

Leaflet provides a number of basemaps for mapping. If you add map tiles using `addTiles()`, you'll get the open street map basemap. Other basemaps can be added using `addProviderTiles()`. The names of available options can be found using:

```{r leaflet-names}
names(providers)
```

As you can see, there are a ton of choices! We'll use `CartoDB.Positron` here, but feel free to pick one that you like for assignments if we don't specify what you should use. Make sure that your other cartographic selections, such as color, do not clash with your basemap. 

The basic `leaflet` workflow involves piping functions together (the `%>%` operator). Each time to see the pipe, think of the word "then." For example, the following code chunk would read:

1.Take the `city84` object, **then**
2. use it as the basis for creating a `leaflet` object with `leaflet()`, **then**
3. add a basemap using the `CartoDB.Positron` tiles, **then**
4. add polygons and create a popup.

```{r leaflet-city}
city84 %>%
  leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(popup = ~NAME)
```

We can get more into the weeds with the neighborhood data since they have additional features. We can create more detailed popups using the `base::paste()` function and some html tags. The most important html tags to know are:

* `<b>text</b>` - bold text
* `<em>text</em>` - italicized text
* `<br>` - line break

```{r leaflet-nhoods}
nhoods84 %>%  
  leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(popup = paste("<b>Name:</b> ", nhoods84$NHD_NAME, "<br>",
                            "<b>2017 Population:</b> ", round(nhoods84$pop17, digits = 0)))
```

### Mapping Quantities with `leaflet`

If we want to turn this into a thematic choropleth map, we can add some additional parameters to the `addPolygons()` function. The options included here are:

* `color` - outline ("stroke") color for each polygon
* `weight` - stroke width
* `opacity` - stroke opacity
* `smoothFactor` - allows `leaflet` to simplify polygons depending on zoom
* `fillOpacity` - fill opacity
* `fillColor` - creates the fill itself
* `highlightOptions` - creates effect when mouse drags over specific polygons

What I have here are good default settings for most of these options, but feel free to experiment!

When we created our popup, we want to round our values so that we don't see the very long real number associated with our data. By using `base::round(var, digits = 0)`, we round to the nearest integer. `digits = 2` would give us two decimal places in contrast.

```{r leaflet-nhoods3}
# create color palette
npal <- colorNumeric("YlOrRd", nhoods84$pop17)

# create leaflet object
nhoods84 %>%
  leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(
    color = "#444444", 
    weight = 1, 
    opacity = 1.0, 
    smoothFactor = 0.5,
    fillOpacity = 0.5,
    fillColor = ~npal(pop17),
    highlightOptions = highlightOptions(color = "white", weight = 2, bringToFront = TRUE),
    popup = paste("<b>Name:</b> ", nhoods84$NHD_NAME, "<br>",
                  "<b>2017 Population:</b> ", round(nhoods84$pop17, digits = 0))) 
```

Next, we should add a legend to make the map easier to interpret. This is done with the `addLegend()` argument. The `opacity` argument in `addLegend()` should match the `fillOpacity` argument in `addPolygons()`!

```{r leaflet-nhoods4}
# create color palette
npal <- colorNumeric("YlOrRd", nhoods84$pop17)

# create map
nhoods84 %>%
  leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(
    color = "#444444", 
    weight = 1, 
    opacity = 1.0, 
    smoothFactor = 0.5,
    fillOpacity = 0.5,
    fillColor = ~npal(pop17),
    highlightOptions = highlightOptions(color = "white", weight = 2, bringToFront = TRUE),
    popup = paste("<b>Name:</b> ", nhoods84$NHD_NAME, "<br>",
                  "<b>2017 Population:</b> ", round(nhoods84$pop17, digits = 0))) %>%
    addLegend(pal = npal, values = ~pop17, opacity = .5, title = "Population (2017)")
```

The colors come from the `RColorBrewer` package. We can use `RColorBrewer::display.brewer.all()` to identify other color ramps:

```{r display-brewer}
display.brewer.all(type = "seq")
```

We'll try the `YlGnBu` palette on this next map, which normalizes the data but dividing our `AREA` variable (which is the area of each neighborhood in square meters) by `1000000`to convert to square kilometers. We use the `/` mathematical operator to divide our values, and parentheses to ensure order of operations is respected. We add a corresponding value to our popup as well:

```{r leaflet-nhoods5}
# create color palette
npal <- colorNumeric("YlGnBu", nhoods84$pop17/(nhoods84$AREA/1000000))

# create map
nhoods84 %>%
  leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(
    color = "#444444", 
    weight = 1, 
    opacity = 1.0, 
    smoothFactor = 0.5,
    fillOpacity = 0.5,
    fillColor = ~npal(pop17/(AREA/1000000)),
    highlightOptions = highlightOptions(color = "white", weight = 2, bringToFront = TRUE),
    popup = paste("<b>Name:</b> ", nhoods84$NHD_NAME, "<br>",
                  "<b>2017 Population:</b> ", round(nhoods84$pop17, digits = 0), "<br>",
                  "<b>2017 Population per Square Kilometer:</b> ", 
                      round((nhoods84$pop17/(nhoods84$AREA/1000000)), digits = 0))) %>%
    addLegend(pal = npal, values = ~pop17/(AREA/1000000), opacity = .5, 
              title = "Population Density (2017)")
```

For our final `leaflet` map, we'll change the color ramp again and update the map to focus on 1950 instead of 2017:

```{r leaflet-nhoods6}
# create color palette
npal <- colorNumeric("OrRd", nhoods84$pop50/(nhoods84$AREA/1000000))

# create map
nhoods84 %>%
  leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(
    color = "#444444", 
    weight = 1, 
    smoothFactor = 0.5,
    opacity = 1.0, 
    fillOpacity = 0.5,
    fillColor = ~npal(pop50/(AREA/1000000)),
    highlightOptions = highlightOptions(color = "white", weight = 2, bringToFront = TRUE),
    popup = paste("<b>Name:</b> ", nhoods84$NHD_NAME, "<br>",
                  "<b>1950 Population:</b> ", round(nhoods84$pop50, digits = 0), "<br>",
                  "<b>1950 Population per Square Kilometer:</b> ", 
                      round((nhoods84$pop50/(nhoods84$AREA/1000000)), digits = 0))) %>%
    addLegend(pal = npal, values = ~pop50/(AREA/1000000), opacity = .5, 
              title = "Population Density (1950)")
```



```{r move-to-docs, include=FALSE}
# you do need to include this in any notebook you create for this class
fs::file_copy(here::here("examples", "lecture-05.nb.html"), 
              here::here("docs", "index.nb.html"), 
              overwrite = TRUE)
```
