---
title: "Lab-13 Replication"
author: "Branson Fox"
date: '(`r format(Sys.time(), "%B %d, %Y")`)'
output:
  github_document: default
  html_notebook: default
---

## Introduction
This notebook replicates Lab-13 - Interactive Mapping of Structures at Risk.

## Dependencies
This notebook requires the packages we used and discussed in lecture-05.

```{r load-packages}
# spatial packages
library(leaflet)      # interactive maps
library(mapview)      # preview spatial data
library(sf)           # tools for spatial data

# other packages
library(here)         # file path management
library(RColorBrewer) # color palettes
```

## Load Data
This notebook requires the data stored in `data/`. Remember that we use `st_read()` to read shapefiles, and `here()` to specify where these files are located.

```{r load-data}
mobileHomes <- st_read(here("data", "METRO_STRUCTURE_PctMobileHome", "METRO_STRUCTURE_PctMobileHome.shp"),
                       stringsAsFactors = FALSE)
```

## Part 1: Data Exploration
### Question 1
We use the `str()` function to list the variables in our data.

```{r p1-q1}
str(mobileHomes)
```

Some of these variables may seem confusing, but the more you work with spatial data, they become intuitive. For example `STATEFP` is the state FIPS code. Missouri is 29 and Illinois is 17. The important variable in these data is `MobileHome` which estimates the percent of homes that are mobile homes.

### Question 2
Now, we will create an interactive preview of the data using the `mapview()` function.

```{r p1-q2}
mapView(mobileHomes)
```

We now have an interactive preview of the data.

## Part 2: Interactive Mapping
### Question 3
Now, we will use leaflet to create an interactive map. But first, we should use the following function to choose a basemap.

```{r p2-q3}
names(providers)
```

For the replication, I'll use `Stamen.TonerLite`. If you would like to preview the basemaps provided by leaflet, you can do so [here](https://leaflet-extras.github.io/leaflet-providers/preview/).

### Question 4
Now, we will use leaflet to symbolize the percentage of mobile homes in each county. We will discuss projections more in depth later, but for now, we will just know that leaflet requires our data to be in `WGS 1984`. To re-project our data, we use the `st_transform()` function:

```{r p2-transform}
mobileHomes84 <- st_transform(mobileHomes, crs = 4326)
```

With our data are now re-projected, we'll make our map:

```{r p2-q4}
# first, define a color ramp
ramp <- colorNumeric("YlGnBu", mobileHomes84$MobileHome)

# then, we will add it to our leaflet code
mobileHomes84 %>%
  leaflet() %>%
  addProviderTiles(providers$Stamen.TonerLite) %>%
  addPolygons(
    color = "#444444",
    weight = 1,
    opacity = 1.0,
    smoothFactor = 0.5,
    fillOpacity = 0.5,
    fillColor = ~ramp(MobileHome),
    highlightOptions = highlightOptions(color = "white", weight = 2, bringToFront = TRUE),
    popup = paste("<b>County:</b> ", mobileHomes84$NAME, "<br>",
                  "<b>Percentage of Mobile Homes:</b> ", mobileHomes84$MobileHome)) %>%
  addLegend(pal = ramp, values = ~MobileHome, opacity = .5, title = "Percentage of Mobile Homes")
```

## Appendix
If you were going to work through this iteratively, here are some logical ways to start with a simple map and make it more complex. First, create a simple interactive leaflet map:

```{r p2-leaflet}
mobileHomes84 %>%
  leaflet() %>%
  addProviderTiles(providers$Stamen.TonerLite) %>%
  addPolygons()
```

Now, we will add additional formatting to our leaflet code to symbolize mobile homes:

```{r p2-leaflet2}
# first, define a color ramp
ramp <- colorNumeric("YlGnBu", mobileHomes84$MobileHome)

# then, we will add it to our leaflet code
mobileHomes84 %>%
  leaflet() %>%
  addProviderTiles(providers$Stamen.TonerLite) %>%
  addPolygons(
    color = "#444444",
    weight = 1,
    opacity = 1.0,
    smoothFactor = 0.5,
    fillOpacity = 0.5,
    fillColor = ~ramp(MobileHome),
    highlightOptions = highlightOptions(color = "white", weight = 2, bringToFront = TRUE))
```

Next, we will add pop-up labels:

```{r p2-leaflet3}
# first, define a color ramp
ramp <- colorNumeric("YlGnBu", mobileHomes84$MobileHome)

# then, we will add it to our leaflet code
mobileHomes84 %>%
  leaflet() %>%
  addProviderTiles(providers$Stamen.TonerLite) %>%
  addPolygons(
    color = "#444444",
    weight = 1,
    opacity = 1.0,
    smoothFactor = 0.5,
    fillOpacity = 0.5,
    fillColor = ~ramp(MobileHome),
    highlightOptions = highlightOptions(color = "white", weight = 2, bringToFront = TRUE),
    popup = paste("<b>County:</b> ", mobileHomes84$NAME, "<br>",
                  "<b>Percentage of Mobile Homes:</b> ", mobileHomes84$MobileHome))
```

And finally we will add a legend to complete our interactive map:

```{r leaflet4}
# first, define a color ramp
ramp <- colorNumeric("YlGnBu", mobileHomes84$MobileHome)

# then, we will add it to our leaflet code
mobileHomes84 %>%
  leaflet() %>%
  addProviderTiles(providers$Stamen.TonerLite) %>%
  addPolygons(
    color = "#444444",
    weight = 1,
    opacity = 1.0,
    smoothFactor = 0.5,
    fillOpacity = 0.5,
    fillColor = ~ramp(MobileHome),
    highlightOptions = highlightOptions(color = "white", weight = 2, bringToFront = TRUE),
    popup = paste("<b>County:</b> ", mobileHomes84$NAME, "<br>",
                  "<b>Percentage of Mobile Homes:</b> ", mobileHomes84$MobileHome)) %>%
  addLegend(pal = ramp, values = ~MobileHome, opacity = .5, title = "Percentage of Mobile Homes")
```
